---
title: "Negative Binomial Mixed Model"
subtitle: 'Example analysis'
author: "Developed by [Gabriel Hoffman](http://gabrielhoffman.github.io/)"
date: "Run on `r format(Sys.time())`"
documentclass: article
vignette: >
  %\VignetteIndexEntry{Negative Binomial Mixed Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\usepackage[utf8]{inputenc}
---



<!---

cd /Users/gabrielhoffman/workspace/repos/fastglmm/vignettes

# rm -rf nbmm_cache/

rmarkdown::render("nbmm.Rmd")

--->


<style>
body {
text-align: justify}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  error = FALSE,
  tidy = FALSE,
  dev = c("png"),
  cache = TRUE
)
```

Recent advances in single cell transcriptomics have enabled generation of large-scale datasets of millions of cells from hundreds or thousands of samples [(Rood, et al., 2024)](https://doi.org/10.1038/s41586-024-08338-4).  In differential expression analysis, the goal is to identify genes whose expression is associated with a variable of interest.  Single cell transcriptomics produces counts for each gene and each observed cell.  

To be concrete, consider counts from the gene PTPRG from 249 subjects measured across 60k microglia cells where an average of 202 cells were observed per subject [(Hoffman, et al, 2025)]( https://doi.org/10.1101/2023.03.17.533005).  Consider the dataset `PsychAD` storing observed cells as rows where column `PTPRG` is the number of counts for this gene, `libSize` is the total counts for each cell, `Dx` is Alzheimer's disease status, `Age` and `Sex` indicate subject age and sex, and `SubID` indicates which subject each observed cell is from.  Here we use a negative binomial mixed model (NBMM) to test if the counts of the PTPRG gene are associated with Alzheimer's disease status after accounting for the total number of counts for each cell.  Age and sex are included as covariates and subject is a random effect to account for the repeated measures design. We use the familiar formula syntax, set `log(libSize)` to be an `offset` term, and use the `fastglmm.nb()` function:

# Fit NBMM
```{r nbmm}
library(fastglmm)

data(PsychAD)

form <- PTPRG ~ offset(log(libSize)) + Dx + Age + Sex + (1|SubID)
fit <- fastglmm.nb(form, PsychAD)

fit
```


# Hypothesis testing
```{r summary}
summary(fit)
```

The PTPRG gene has increased expression in subjects with Alzheimer's disease compared with controls with log fold change $1.102755$  and a p-value $< 2\times 10^{-16}$.
The count data shows high overdispersion with $\theta = 0.2443$, where $\theta = \infty$ is equivalent to a Poisson distribution with no overdispersion.

Note that while 71\% of the 60k observations are zero, this is due to a low count rate rather than zero inflation.  This is supported by comparison with the zero inflation component using `glmmTMB::glmmTMB()` (p-value of zero-inflation intercept: 0.961), and consistent with other work on single cell transcriptomics  ([Jiang, et al., 2022](https://doi.org/10.1186/s13059-022-02601-5), [Sarkar and Stephens, 2021](https://doi.org/10.1038/s41588-021-00873-4), [Svensson, 2020](https://doi.org/10.1038/s41587-019-0379-5)).


# Variance partitioning analysis
```{r varpart}
varpart(fit)
```
Variance partitioning analysis quantifies the contribution of each variable to the observed variance in PTPRG expression.  Alzheimer's disease status explains 3.8% of observed expression variation, variation across subjects explains 6.0%, with sex and age making a smaller contribution.  Measurement error  due to sampling a finite number of counts (i.e. `CountNoise`) explains 55.2% of the variance, and the remaining 34.8% of the variance is explained by the residuals.

#### Session info
<details>
```{r sessionInfo}
sessionInfo()
```
</details>
