---
title: "Mediation analysis with fastglmm"
subtitle: 'Simulation study'
author: "Developed by [Gabriel Hoffman](http://gabrielhoffman.github.io/)"
date: "Run on `r format(Sys.time())`"
documentclass: article
vignette: >
  %\VignetteIndexEntry{Mediation analysis with fastglmm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\usepackage[utf8]{inputenc}
---



<!---

rmarkdown::render("mediation_simulation.Rmd")

--->


<style>
body {
text-align: justify}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  error = FALSE,
  tidy = FALSE,
  dev = c("png"),
  cache = TRUE
)
```

```{r load, cache=FALSE}
library(mediation)
library(fastglmm)
library(parallel)
library(PRROC)
library(ggplot2)
```


# Gaussian
```{r sim}
# Model
# x = a*z + noise
# y = b*x + c_prime*z + noise

# ADE is c_prime
# ACME is a*b

a = 3
b = 1
c_prime = -1
sig_x = 3
sig_y = 2

n <- 100
n_sims = 100

res = mclapply(seq(n_sims), function(i){

  c_prime.sim = ifelse(i < n_sims / 2, c_prime, 0)

  data <- data.frame(z = rnorm(n))
  data$x <- with(data, a*z + rnorm(n,0,sig_x))
  data$y <- with(data, b*x + c_prime.sim*z + rnorm(n,0,sig_y))
  data$letter <- sample(LETTERS[1:2], n, replace=TRUE)

  fit1 <- glm(x ~ z, data=data)
  fit2 <- glm(y ~ x + z, data=data)

  res <- mediation:::mediate(fit1, fit2, treat="z", mediator="x")

  df <- summary(res)

  data.frame(i = i, 
            c_prime = c_prime.sim,
            ADE.hat = df$z.avg, 
            ADE.p = df$z.avg.p)
}, mc.cores=6)
res = do.call(rbind, res)

idx = res$c_prime == 0
fg = -log10(res$ADE.p[!idx] + 1e-20)
bg = -log10(res$ADE.p[idx]+ 1e-20)
pr <- pr.curve(scores.class0 = fg, scores.class1 = bg, curve=T)
plot(pr)

 pr$auc.davis.goadrich
```

# Poisson
```{r sim.poisson}
# Model
# x = a*z + noise
# y = b*x + c_prime*z + noise

# ADE is c_prime
# ACME is a*b

mu.x = 2
mu.y = 3
a = 1
b = 1
c_prime = -1
sig_x = .1
sig_y = .1

n <- 2000
n_sims = 100

res = mclapply(seq(n_sims), function(i){

  c_prime.sim = ifelse(i < n_sims / 2, c_prime, 0)

  data <- data.frame(z = rnorm(n))

  data$eta.x <- with(data, mu.x + a*z)
  data$x <- with(data, eta.x + rnorm(n,0,sig_x))
  # data$x <- with(data, rpois(n, exp(eta.x)))

  data$eta.y <- with(data, mu.y + b*x + c_prime.sim*z)
  # data$y <- with(data, eta.y + rnorm(n,0,sig_y))
  data$y <- with(data, rpois(n, exp(eta.y)))

  data$letter <- sample(LETTERS[1:2], n, replace=TRUE)

  fit1 <- fastglmm(x ~ z + (1|letter), data=data)
  fit2 <- fastglmm(y ~ x + z + (1|letter), data=data, family=poisson())
  # fit2 <- fastlmm(log(y) ~ x + z + (1|letter), data=data)

  df <- fastglmm::mediate(fit1, fit2, treat="z", mediator="x",)


  fit1 <- glm(x ~ z, data=data)
  fit2 <- glm(y ~ x + z, data=data, family=poisson())





  data.frame(i = i, 
            c_prime = c_prime.sim,
            ADE.hat = df$Estimate[2], 
            ADE.p = df$p.value[2])
}, mc.cores=8)
res = do.call(rbind, res)

idx = res$c_prime == 0
fg = -log10(res$ADE.p[!idx] + 1e-20)
bg = -log10(res$ADE.p[idx]+ 1e-20)
pr <- pr.curve(scores.class0 = fg, scores.class1 = bg, curve=T)
plot(pr)

plot(res$c_prime, res$ADE.hat)
```


```{r sim2}


n = 2000
mu.x = 3
mu.y = 4
a = 2
b = 1
c_prime = 1
sig_x = 4
sig_y = .1

c_prime.sim = c_prime

data <- data.frame(z = rnorm(n))

data$eta.x <- with(data, mu.x + a*z)
data$x.l <- data$eta.x + rnorm(n,0,sig_x)
data$x <- rpois(n, exp(data$x.l))

data$y.l <- with(data, log(mu.y + b*exp(x.l) + c_prime.sim*z))
data$y <- rpois(n, exp(data$y.l))

fit1 <- glm(x ~ z, data=data, family=poisson())
data$x.g <- log(data$x+1)
fit2 <- glm(y ~ x.g + z, data=data, family=poisson())

coef(fit2)
```



```{r sim3}
library(parallel)
library(PRROC)

n = 2000
mu.x = 4
mu.y = 4
a = .01
b = .01
c_prime = .3
sig_x = 1
sig_y = .1

n_sims = 10000

res = mclapply(seq(n_sims), function(i){

  b.in = ifelse(i <= n_sims / 2, b, 0)

  # simulate data
  data <- data.frame(z = rnorm(n))
  data$eta.x <- mu.x + with(data, a*z)  + rnorm(n,0,sig_x)
  data$x <- rpois(n, exp(data$eta.x))
  data$eta.y = with(data, b.in*eta.x + c_prime*z)
  data$y <- rpois(n, exp(data$eta.y + mu.y))

  res <- twoStageGLM("z", "x", "y", poisson(), poisson(), data)


  # fit.m = glm(x ~ z, data, family=poisson())
  # fit.y = glm(y ~ x + z, data, family=poisson())
  # res.med = mediation::mediate(fit.m, fit.y, treat="z", mediator="x")

  df = res$coef
  # df$mediate.p = res.med$d0.p
  df$b.true = b.in
  df$ME.true = a*b.in
  df

}, mc.cores=8)
res = do.call(rbind, res)

idx = res$b.true == 0
fg = -log10(res$ME.p[!idx])
bg = -log10(res$ME.p[idx])
pr <- pr.curve(scores.class0 = fg, scores.class1 = bg, curve=T)
plot(pr)






hist(res$ME.p[idx])

boxplot(ME ~ ME.true, res)


```



library(sadists)

mu = c(2, 1)
sig = c(1, 2)

x1 <- rnorm(1e6, mu[1], sig[1])
x2 <- rnorm(1e6, mu[2], sig[2])
x  <- x1*x2
d  <- density(x)
plot(d,lwd=2)

z  <- par('usr')
dn <- dnorm(d$x, mean=mean(x), sd=sd(x))
z  <- seq(z[1],z[2],length.out=length(dn))
lines(z, dn, col=2, lwd=2)

d2 = dprodnormal(d$x, mu, sig)
lines(z, d2, col=3, lwd=2, lty=2)

legend('topright', legend=c('Estimated density', 'Normal distribution', "dprodnormal"), lwd=2, lty=c(1,1, 2),col=c(1,2, 3))


plot(ecdf(x), lwd=2)
xval = seq(-40, 40, by=.01)

d2 = pprodnormal(xval, mu, sig)
lines(xval, d2, col=3, lwd=2, lty=2)

y = sapply(xval, function(v){
  pprodnormalMeeker(v, mu[1], mu[2], sig[1], sig[2], lower.tail=TRUE)$p
  })
lines(xval, y, col=2, lwd=2, lty=2)




plot(data$eta.x, predict(fit1))

summary(glm.nb(x ~ z, data=data))


model.m = fit1
model.y = fit2

treat = "z"
mediator = "x"


treatmeant <- "z"
mediator <- "x"
outcome <- "y"
covariates <- "1"
family.mediator <- poisson()
family.outcome <- poisson()


res <- two_stage_glm("z", "x", "y", poisson(), poisson(), data)
res$coef


res.sim = lapply( seq(200), function(i){

  # simuations
  # eta.x = model.matrix(res$fit.mediator) %*% coef(res$fit.mediator) + rnorm(nrow(data), 0, sd=sig_x)
  # # sd(residuals(res$fit.mediator, "work"))
  # x.sim = rpois(nrow(data), exp(eta.x))

  # cor(log(x.sim + .25), data$z)

  # res$fit.outcome$data$x = eta.x 
  # y.sim = simulate(res$fit.outcome)[,1]

  # data.sim = data.frame(z = data$z,
  #                       x = x.sim,
  #                       y = y.sim)

  # data.sim = data.frame(cbind(data.sim, data[,all.vars(covariates),drop=FALSE]))

  data.sim = data.frame(z = data$z)
  a.hat = res$coef$a
  mu.x.hat = coef(res$fit.mediator)["(Intercept)"]
  data.sim$eta.x <- with(data.sim, a.hat*z) + rnorm(n,0, sig_x)
  data.sim$x <- rpois(n, exp(data.sim$eta.x + mu.x.hat))

  b.hat = res$coef$b
  c_prime.hat = res$coef$DE
  mu.y.hat = coef(res$fit.outcome)["(Intercept)"]
  data.sim$eta.y = with(data.sim, b*eta.x + c_prime.sim*z)
  data.sim$y <- rpois(n, exp(data.sim$eta.y + mu.y.hat))

  res2 = two_stage_glm("z", "x", "y", poisson(), poisson(), data.sim)
  res2$coef


})
res.sim = do.call("rbind", res.sim)

colMeans(res.sim)
apply(res.sim, 2, sd)

mean(abs(res.sim$b))


# In simulations, g(x) and z are _strongly_ correlated

# USE simulation code:

data$eta.x <- with(data, a*z)  + rnorm(n,0,sig_x)
data$x <- rpois(n, exp(data$eta.x + mu.x))

data$eta.y = with(data, mu.y + b*eta.x + c_prime.sim*z)
data$y <- rpois(n, exp(data$eta.y))



<!---



rmarkdown::render("mediation_simulation.Rmd")



--->






<!---

  fit2 <- glm(y.obs ~ x + z, data=data, family=poisson())
  fit3 <- lm(log(y.obs) ~ x + z, data=data)


  plot(rowMeans(log(simulate(fit2, 1000))), 
      rowMeans(simulate(fit3, 1000)))

  plot(rowSds(log(as.matrix(simulate(fit2, 1000)))), 
      rowSds(as.matrix(simulate(fit3, 1000))))



  fit1 <- glm(x ~ z, data=data)
  fit2 <- glm(y.obs ~ x + z, data=data, family=poisson())
  res <- mediation:::mediate(fit1, fit2, treat="z", mediator="x")
  summary(res)

  mean(c(res$z0, res$z1))


--->




